---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r loaddata}
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", "./zipped_data")
unzip("./zipped_data")
activity = read.csv("activity.csv")
activity$date <- as.Date(as.character(activity$date), "%Y-%m-%d")
```


## What is mean total number of steps taken per day?
```{r steps}
total_steps_per_day <- tapply(activity$steps, activity$date, sum)
hist(total_steps_per_day, xlab = "Total number of steps per day", main = "Total number of steps per day")
mean_total_steps <- mean(total_steps_per_day, na.rm=TRUE)
median_total_steps <- median(total_steps_per_day, na.rm = TRUE)
```

The mean of total number of steps is `r mean_total_steps` and the median is `r median_total_steps`.



## What is the average daily activity pattern?
```{r average_steps}
avg_steps_per_interval <- tapply(activity$steps, activity$interval, mean, na.rm=TRUE)
plot(x=unique(activity$interval), y=avg_steps_per_interval, type="l", xlab = "Interval", ylab = "Average steps taken in the interval across days", main="Average steps taken in each interval, across days")
index_max <- which.max(avg_steps_per_interval)
max_steps_interval <- unique(activity$interval)[index_max]
```

The interval with maximum average steps across days is `r max_steps_interval`.



## Imputing missing values
```{r missing}
num_NA <- sum(is.na(activity$steps))
```

The number of rows with missing value is `r num_NA`.

These missing values can be substituted using the average number of steps taken during that interval across days.

```{r imputing}
imputed_values <- data.frame(interval=unique(activity$interval), Avg_steps = avg_steps_per_interval)
library(dplyr)
imputed_activity_data <- merge(activity, imputed_values, by="interval")
imputed_activity_data$steps[is.na(imputed_activity_data$steps)] <- imputed_activity_data$Avg_steps[is.na(imputed_activity_data$steps)]
new_total_steps_per_day <- tapply(imputed_activity_data$steps, imputed_activity_data$date, sum)
hist(new_total_steps_per_day, xlab = "Total number of steps per day", main = "Total number of steps per day")
new_mean_total_steps <- mean(new_total_steps_per_day, na.rm=TRUE)
new_median_total_steps <- median(new_total_steps_per_day, na.rm = TRUE)
```
The new mean of total steps taken is `r new_mean_total_steps` and the new median is `r new_median_total_steps`. The numbers are comparable with the original mean and median. Hence the imputation has not introduced any bias.


## Are there differences in activity patterns between weekdays and weekends?
```{r}
activity$type <- ifelse(weekdays(activity$date)=="Saturday" | weekdays(activity$date)=="Sunday", "weekend", "weekday")
weekend_data <- subset(activity, type=="weekend")
weekday_data <- subset(activity, type=="weekday")
avg_steps_weekend <- tapply(weekend_data$steps, weekend_data$interval, mean, na.rm=TRUE)
avg_steps_weekday <- tapply(weekday_data$steps, weekday_data$interval, mean, na.rm=TRUE)
par(mfrow=c(2,1), mar=c(3,2,2,1))
plot(x=unique(activity$interval), y=avg_steps_weekend, type="l", xlab = "Interval", ylab = "Average steps", main="Average steps taken in each interval during weekend")
plot(x=unique(activity$interval), y=avg_steps_weekday, type="l", xlab = "Interval", ylab = "Average steps", main="Average steps taken in each interval during weekday")
```
